Composable Concurrency Models

### 2016-10-27 paper/paper.tex
Finished OBMOP section

Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}. Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable. For example, the atomic operation \emph{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}. \emph{compare-and-swap} ensures that a new value cannot be written based on outdated information. Suppose that a thread $A$ reads a variable $x$ and begins computing a new value. At the same time, another thread $B$ modifies $x$. When $A$ tries to set the value of $x$ via \emph{compare-and-swap}, the write will fail. As a result, all writes on $x$ are guaranteed to be linearizable.

Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible. \emph{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.

@@

Futures and promises are two popular proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual''). Generally, futures are resolved to the result of the completed operation. The result is then accessed implicitly; any use of the future will return its value. Promises, on the other hand, are created as independent objects and require the result to be accessed explicitly.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine (VM) level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt found that virtual machines at the time did not provide intrinsic support for more than one specific model~\cite{Marr2009}.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. To identify concepts relevant for a multi-language VM, Marr and D'Hondt selected four questions~\cite{Marr2012}:

\begin{enumerate}
  \item Can the concept be reasonably implemented as a library? (A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant VM inclusion.)

  \item Does the concept require runtime support to guarantee its semantics? (Semantic guarantees may be enforced by a compiler but not the VM.)

  \item Would runtime support significantly benefit performance? (Deeper knowledge of the concept may enable the VM to better optimize.)

  \item Is the concept already supported by a common VM like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
\end{enumerate}

From those questions, 26 concepts were identified that would benefit from better low-level support.

>>>>

Atomic variables are used to share independent objects that do not demand coordinated updates but need to be shared by multiple threads~\cite{Swalens2014}. Atomic variables can only be read and mutated by operations that are guaranteed to be linearizable. For example, the atomic operation \code{compare-and-swap} compares the current value of a variable to a given value and only performs a write if the values match~\cite{Swalens2014}. \code{compare-and-swap} ensures that a new value cannot be written based on outdated information. Suppose that a thread $A$ reads a variable $x$ and begins computing a new value. At the same time, another thread $B$ modifies $x$. When $A$ tries to set the value of $x$ via \code{compare-and-swap}, the write will fail. As a result, all writes on $x$ are guaranteed to be linearizable.

Atomic operations are often implemented in hardware and guarantee linearizability at the lowest level possible. \code{fetch-and-add}, another atomic operation, is a common processor instruction that executes multiple hardware-level operations: The value old value is copied from a location in memory into a temporary register, addition is performed on the value in the temorary register, and the new value is stored at the original location.

@@

Futures and promises are two of the most common proxy models, though the terms are frequently used interchangeably (along with ``delayed,'' ``deferred,'' and ``eventual''). Generally, futures are resolved to the result of the completed operation. The result is then accessed implicitly; any use of the future will return its value. Promises, on the other hand, are created as independent objects and require the result to be accessed explicitly.

@@

Marr and D'Hondt~\cite{Marr2012} surveyed a variety of concurrent and parallel programming concepts to identify a unifying concurrency abstraction that could be implemented at the virtual machine level.\footnote{A 2009 survey by Marr, Haupt, and D'Hondt~\cite{Marr2009} found that virtual machines at the time did not provide intrinsic support for more than one specific model.} A unifying abstraction would enable each concurrency model in a high-level language to be represented in machine code without sacrificing performance or semantics. To identify concepts relevant for a multi-language virtual machine, Marr and D'Hondt selected four questions:

\begin{enumerate}
  \item Can the concept be reasonably implemented as a library? (A library implementation may suffer from loss of performance or semantics to an extent that the concept would warrant virtual machine inclusion.)

  \item Does the concept require runtime support to guarantee its semantics? (Semantic guarantees may be enforced by a compiler but not the virtual machine. For instance, immutable objects in one language may not be immutable in machine code, and therefore may be mutated by another language.)

  \item Would runtime support benefit performance? (Deeper knowledge of the concept may enable the virtual machine to better optimize. Some concepts, though, may require knowledge of the underlying hardware to realize any performance improvements; such concepts would not necessarily benefit from runtime support.)

  \item Is the concept already supported by a common virtual machine like the Java Virtual Machine (JVM) or the Common Language Runtime (CLR)?
\end{enumerate}

From those questions, 26 concepts were identified that would benefit from better low-level support. Of those, 18 suffered from loss of semantic guarantees when compiled to machine code. With that in mind, Marr and D'Hondt derived the following requirements for a unifying abstraction:

\begin{itemize}
  \item Managed mutation: Concurrency models impose restrictions on how state can be modified; therefore, mutation must be able to be flexibly managed.

  \item Managed execution: Similarly, invocation of operations on objects is usually restricted and must be able to be flexibly managed.

  \item Ownership: Some owning entity usually regulates mutation and execution, so ownership should be supported in a way that allows for adaptable mutation and execution rules.

  \item Leveled reflection: Reflection allows a program to examine its own structure and behavior at runtime. To guarantee safety, there needs to be a distinction between language-level reflection (restricted reflection operations supported by the high-level language) and meta-level reflection (unrestricted reflection operations on the low-level abstraction).

  \item Enforceability: All defined restrictions should be enforceable on different concurrency models. Additionally, the semantics of the language in which a model is defined must be enforced regardless of where the model is used.
\end{itemize}

\begin{figure}[h]
  \centering
  \includegraphics[width=224pt]{obmop}
  \caption{Marr and D'Hondt's object-based meta-object protocol~\cite{Marr2012}.}
\label{figure:obmop}
\end{figure}

Given those requirements, Marr and D'Hondt defined an ownership-based meta-object protocol (MOP) that could describe a low-level abstraction, illustrated in Figure~\ref{figure:obmop}. The owner of an object, referred to as the domain, manages operations on all of the objects that it owns (namely, reading and writing of object fields and invocation of methods on objects). This satisfies the ownership requirement. Read and write operations are directed to the \code{readField} and \code{writeField} operations on the domain, satisfying the managed mutation requirement. Similarly, all method invocations are directed to the \code{exec}, satisfying the managed execution requirement. Threads are executed in a domain and specify whether execution occurs at the language level (with restricted reflection) or the meta level (with unrestricted reflection), satisfying the leveled reflection requirement. Globally shared resources are handled by the domain (specifically, the \code{readGlobal} and \code{writeGlobal} methods) if they might break semantics. Additionally, the \code{primitive*} operations allow callers to override the semantics of virtual machine primitives, completing the enforcability requirement.

To demonstrate the suitability of the MOP as a low-level abstraction, Marr and D'Hondt implemented several concurrency models. The MOP was able to support LRSTM (a STM for the Smalltalk language), Clojure's agents (which rely on communicating threads), active objects (a variation on proxies), and several other models. While the MOP was able to successfully enforce the semantics of each model, the resulting performance cost suggested that the MOP must be implemented at the virtual machine level to sufficiently reduce the abstraction overhead. Marr and D'Hondt also note that regarding the owner of an object as the only entity able to restrict interaction is somewhat limiting. For instance, the behavior of the system when different semantics interact needs to be explicitly specified.
